---
title: "class14_Covid_Mini-Project"
author: "Seong Tae Gwon"
date: "3/7/2022"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

# Background

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

# Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv") 
head(vax)
```

> Q1. What column details the total number of people fully vaccinated?

#### Answer:
The column `persons_fully_vaccinated` details the total number of people fully vaccinated.

> Q2. What column details the Zip code tabulation area? 

#### Answer:
The column `zip_code_tabulation_area` details the Zip code tabulation area.

> Q3. What is the earliest date in this dataset? 

```{r}
head(vax$as_of_date)
```

#### Answer:
The earliest date in this dataset is `2021-01-05` (YYYY-MM-DD).

> Q4. What is the latest date in this dataset?

```{r}
tail(vax$as_of_date)
```

#### Answer:`
As of 3/1/22, the latest date in this dataset is `2022-03-01` (YYYY-MM-DD).

Note: Lab has `2022-02-22` as the lastest date. 

Let’s call the skim() function from the skimr package to get a quick overview of this dataset.

```{r}
#install.packages("devtools")
#devtools::install_github("ropensci/skimr")
library(skimr)
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

#### Answer:
As of 3/1/22, there are 10 numeric columns in this dataset.

Note: Lab has 9 numeric columns as answer. 

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

#### Answer:
As of 3/1/22, there are 18,338 NA values in the `persons_fully_vaccinated` column.

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax)
```

#### Answer:
As of 3/1/22, 17.04% of `persons_fully_vaccinated` values are missing.

> Q8. [Optional]: Why might this data be missing?

#### Answer:
Individuals currently in US vaccine clinical trials are considered unvaccinated until the trial has concluded (2 weeks). 

## Working with dates

One of the “character” columns of the data is as_of_date, which contains dates in the Year-Month-Day format.

Dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier.

```{r}
#install.packages("lubridate")
library(lubridate)
```

What is today's date?

```{r}
today()
```

The as_of_date column of our data is currently not that usable. For example, we can’t easily do math with it like answering the simple question of how many days have passed since data was first recorded.

However, if we convert our date data into a lubridate format, this will be much easier (as well as plotting time series data later on).

```{r}
 # Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]
```

Now we can do math with dates. How many days have passed since the first vaccination reported in this dataset?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

#### Answer:
It has been 6 days since the last update of this dataset.

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

#### Answer:
There are 61 unique dates in this dataset.

# Working with ZIP codes

One of the numeric columns in the dataset (namely vax$zip_code_tabulation_area) are actually ZIP codes - a postal code used by the United States Postal Service (USPS). In R we can use the zipcodeR package to make working with these codes easier. For example, let’s install and then load up this package and to find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code area.

```{r}
#install.packages("zipcodeR")
library(zipcodeR)
# Centroid of LA Jolla 92037 ZIP code area
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92109") )
```

We can use this reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.

```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

## Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county == "San Diego", ]
```

```{r}
# Using dplyr code
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

#### Answer:
There are 107 distinct zip codes listed for San Diego County.

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

#### Answer:
ZIP code 92154 has the largest 12+ population.

Using dplyr select all San Diego “county” entries on “as_of_date” “2022-02-22” and use this for the following questions.

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”?

```{r}
 sd.percentfc <- filter(vax, county == "San Diego" & as_of_date == "2022-02-22")
mean(sd.percentfc$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

#### Answer:
The overall average “Percent of Population Fully Vaccinated” value is 0.7042.

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?

#### Answer:
```{r}
library(ggplot2)

hist(sd.percentfc$percent_of_population_fully_vaccinated,
     main="Histogram of Vaccination Rates Across San Diego County", 
     xlab="Percent Fully Vaccinated on 2022-02-22",
     ylab="Frequency")
```

```{r}
ggplot(sd.percentfc) + 
  aes(percent_of_population_fully_vaccinated) + 
  geom_histogram(bins = 12) +
  labs(x="Percent of Population Fully Vaccinated in a Zip Code Area",
       y="Count (ZIP code areas)", 
       title="Histogram of Vaccination Rates Across San Diego County",
       subtitle="As of 2022-02-22")
```

## Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot, make a graph of the vaccination rate time course for the 92037 ZIP code area.

#### Answer:
```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

## Comparing to similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2022-02-22”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

#head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean.36
```

#### Answer:
The mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22” is 0.7334.

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept=mean.36, col = "red", linetype = "dashed")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

#### Answer:
```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

#### Answer:
```{r}
ggplot(vax.36)+
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=30) +
  labs(x="Percent Vaccinated", y="Count")
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

#### Answer:
Calculated average value for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22” is 0.7334. Both 92040 and 92109 ZIP code areas are below the calculated average value (0.5513 and 0.7230 respectively).

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

#### Answer:
```{r}
 vax.36.all <- filter(vax, age5_plus_population > 36144)
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated,
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0, 1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only populations with a population above 36k are shown.") +
  geom_hline(yintercept = mean.36, linetype= "dashed")
```


> Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

#### Answer:
I don't mind traveling for Spring break since the vaccination rate across CA has been above 75 percentile. However, I still prefer having remote/hybrid class.