---
title: "HW class10 Pt.2"
author: "Seong Tae Gwon"
date: "2/21/2022"
output:
  pdf_document: default
  html_document: default
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble <https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894970-39895222;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel>

Here, we read this CSV file.

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl)*100
```

Now, let's look at a different population. I picked GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G.

```{r}
table(gbr$Genotype..forward.strand.) / nrow(gbr)*100
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.

# Note: Section 2. Initial RNA-Seq analysis and Section 3. Mapping RNA-Seq reads to genome on lab PDF submission

Section 4. Population Scale Analysis

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

How many samples do we have?

```{r}
# Total sample size
nrow(expr)
```


> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
# Sample size for each genotype
table(expr$geno)
```

```{r}
bp <- boxplot(exp~geno, data=expr)
```

```{r}
# Third row contains mdeian values for each genotype
bp$stats
```

#### Answer:
Genotype A|A has sample size of 108 and median value of 31.25.
Genotype A|G has sample size of 233 and median value of 25.06.
Genotype G|G has sample size of 121 and median value of 20.07.

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

# Make a boxplot
ggplot(expr) + aes(x=geno,y=exp,fill=geno) +
  geom_boxplot(notch=TRUE)
```

#### Answer:
Homozygous A|A genotype has higher relative expression than that of G|G (i.e. A|A is up-regulated and G|G is down-regulated). Hence, this indicate that the SNP affect the expression of ORMDL3.